Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-04-15 Paper/manuscript_v2/manuscript_07_01.tex
Fixes in bib
fix variance in equations B6 and B7

\usepackage{listings}
%\usepackage{slashbox}
\usepackage{caption}
\captionsetup{font=footnotesize}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black]{hyperref}

@@

Additionally, there is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the approximated function. Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} depict how these three factors interact with each other in relation to a close approximation of the HSGP model, in the cases of a GP with square exponential covariance function and Mat\'ern($\nu$=3/2) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a squared exponential covariance function) with lengthscale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum $m$ required to achieve a accurate approximation in the sense of satisfying equation (\ref{eq_diff_covs}). Similarly for Figure \ref{fig6_lscale_vs_J_vs_c_Matern} in the case of a Mat\'ern($\nu$=3/2) covariance function. We considered an approximation to be a close enough when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in equation (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$:
%
\begin{equation*}
 \frac{\varepsilon}{\int k(\tau) \,\mathrm{d}\tau} < 0.01.
\end{equation*}
%
Alternatively, these figures could be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be closely approximated given $m$ and $c$. We obtain the following main conclusions:

@@

The function values of a GP model with this low-rank representation of the periodic exponential covariance function can be easily derived. Considering the identity
%
\begin{equation*}
\cos(j\omega_0 (x-x'))=\cos(j\omega_0 x) \cos(j\omega_0 x') + \sin(j\omega_0 x) \sin(j\omega_0 x'),
\end{equation*}
the covariance $k(\tau)$ in equation (\ref{ch5_eq_cov_periodic_taylor_approx}) can be written as
%
\begin{equation} \label{ch5_eq_cov_periodic_taylor_approx_2}
k(x,x')= \sigma^2 \left( \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 x)  \cos(j\omega_0 x') + \sum_{j=1}^{J} \tilde{q}_j^2 \sin(j\omega_0 x) \sin(j\omega_0 x') \right).
\end{equation}
With this approximation for the periodic squared exponential covariance function $k(x,x')$, the approximate GP model $f(x) \sim \GP\left(0,k(x,x')\right)$ equivalently leads to a linear representation of $f(\cdot)$ via
%
\begin{equation} \label{ch5_eq_f_period}
f(x) \approx \sigma \left( \sum_{j=0}^J  \tilde{q}_j \cos(j\omega_0 x) \beta_j +  \sum_{j=1}^J \tilde{q}_j \sin(j\omega_0 x) \beta_{J+1+j} \right),
\end{equation}
where $\beta_j \sim \Normal(0,1)$, with $j=1,\dots,2J+1$. The cosine $\cos(j\omega_0 x)$ and sinus $\sin(j\omega_0 x)$ terms do not depend on the covariance hyperparameters $\ell$. The only dependence on the hyperparameter $\ell$ is through the coefficients $\tilde{q}_j$, which are $J$-dimensional. The computational cost of this approximation scales as $O\big(n(2J+1) + (2J+1)\big)$, where $n$ is the number of observations and $J$ the number of cosine terms.
The parameterization in equation (\ref{ch5_eq_f_period}) is naturally in the non-centered form with independent prior distributions on
$\beta_j$, which makes posterior inference easier.

\section*{Acknowledgment}

>>>>

\usepackage{listings}
%\usepackage{slashbox}
\usepackage{caption}
\captionsetup{font=footnotesize}
%\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black]{hyperref}

@@

Additionally, there is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the approximated function. Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} depict how these three factors interact with each other in relation to a close approximation of the HSGP model, in the cases of a GP with square exponential covariance function and Mat\'ern($\nu$=3/2) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a squared exponential covariance function) with lengthscale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum $m$ required to achieve a accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). Similarly for Figure \ref{fig6_lscale_vs_J_vs_c_Matern} in the case of a Mat\'ern($\nu$=3/2) covariance function. We considered an approximation to be a close enough when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$:
%
\begin{equation*}
 \frac{\varepsilon}{\int k(\tau) \,\mathrm{d}\tau} < 0.01.
\end{equation*}
%
Alternatively, these figures could be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be closely approximated given $m$ and $c$. We obtain the following main conclusions:

@@

The function values of a GP model with this low-rank representation of the periodic exponential covariance function can be easily derived. Considering the identity
%
\begin{equation*}
\cos(j\omega_0 (x-x'))=\cos(j\omega_0 x) \cos(j\omega_0 x') + \sin(j\omega_0 x) \sin(j\omega_0 x'),
\end{equation*}
the covariance $k(\tau)$ in equation (\ref{ch5_eq_cov_periodic_taylor_approx}) can be written as
%
\begin{equation} \label{ch5_eq_cov_periodic_taylor_approx_2}
k(x,x')= \alpha \left( \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 x)  \cos(j\omega_0 x') + \sum_{j=1}^{J} \tilde{q}_j^2 \sin(j\omega_0 x) \sin(j\omega_0 x') \right).
\end{equation}
With this approximation for the periodic squared exponential covariance function $k(x,x')$, the approximate GP model $f(x) \sim \GP\left(0,k(x,x')\right)$ equivalently leads to a linear representation of $f(\cdot)$ via
%
\begin{equation} \label{ch5_eq_f_period}
f(x) \approx \alpha^{1/2} \left( \sum_{j=0}^J  \tilde{q}_j \cos(j\omega_0 x) \beta_j +  \sum_{j=1}^J \tilde{q}_j \sin(j\omega_0 x) \beta_{J+1+j} \right),
\end{equation}
where $\beta_j \sim \Normal(0,1)$, with $j=1,\dots,2J+1$. The cosine $\cos(j\omega_0 x)$ and sinus $\sin(j\omega_0 x)$ terms do not depend on the covariance hyperparameters $\ell$. The only dependence on the hyperparameter $\ell$ is through the coefficients $\tilde{q}_j$, which are $J$-dimensional. The computational cost of this approximation scales as $O\big(n(2J+1) + (2J+1)\big)$, where $n$ is the number of observations and $J$ the number of cosine terms.
The parameterization in equation (\ref{ch5_eq_f_period}) is naturally in the non-centered form with independent prior distributions on
$\beta_j$, which makes posterior inference easier.

%\section*{Acknowledgment}
